Graph Topological Aspects of Granger Causal Network Learning

### 2019-07-09 paper/paper.tex
Some final figures for the eeg analysis.  Should I provide a comparison against LASSO?

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage[backend=bibtex,sorting=none]{biblatex}
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{todonotes}
\bibliography{\string~/Documents/academics/global_academics/global_bib.bib}
% \usepackage{apxproof}

@@

\begin{document}
\maketitle
\abstract{We study Granger-causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.
  We implement and detail finite-sample heuristics derived from our
  theory, and establish through simulation the efficiency gains (both
  statistical and computational) which can be obtained (in comparison
  to LASSO algorithms) when structural assumptions are met.}

\keywords{Granger-causality, causality graph, time series, vector autoregression,
  LASSO, gene regulatory networks, network learning}

@@

\subsection{Contributions}
In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger-causality graph, which to our knowledge remains largely
unexplored.  The focus of this paper is to shed light on some of these
topological questions, in particular, we study a particularly simple
notion of causality graph topology which we term ``strongly causal''
and show that $\VAR$ models having this structure satisfy natural
intuitive notions of ``information flow'' through the graph.
Moreover, we show that such graphs are perfectly recoverable with only
\textit{pairwise} Granger-causality tests, which would otherwise suffer
from serious confounding problems.  Our finite sample results are
based on simulations, where we show promising results for these graphs
topologies where our algorithm performs substantially better in our
simulation setup than do competing LASSO algorithms, even for graphs
that do not exactly satisfy our strongly-causal topology assumptions.

@@

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger-causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger-causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).
Secondly, we propose in section \ref{sec:structure_learning} a graph
search heuristic which implements our theoretical results to finite
data samples, specifying and summarizing appropriate methods for
hypothesis testing (section \ref{sec:pairwise_hypothesis_testing}),
model order selection (section \ref{sec:model_order_selection}),
computationally efficient estimation (section
\ref{sec:efficient_model_estimation}), and error rate controls
(section \ref{sec:error_rate_control}).  Our heuristics are compared
against the adaLASSO algorithm in Section
\ref{sec:empirical_evaluation}.  We stress the scalability of our
algorithm which is capable of comfortably handling hundreds or
thousands of nodes on a single machine, as opposed to standard LASSO
algorithms which do not take advantage of the special structure
associated with stationary time series data.  In section
\ref{sec:application} we develop an example application.  Concluding
remarks on further open problems and extentions are provided in Section
\ref{sec:conclusion}.  \todo{do an application}

% \cite{yuan2006model} as well as local search hill
% climbing (\hl{state here the results}).

@@

    \BlankLine
    \caption{Finite Sample Pairwise Graph Recovery (PWGC)}
    \label{alg:finite_pwgc}
    \Input{Estimates of pairwise Granger-causality statistics $F_{ij}$
      (eqn \ref{eqn:gc_statistics}).  Matrix of edge probabilities $P_{ij}$ (eqn \ref{eqn:edge_inclusion_probability}).  Hypothesis testing threshold $\delta$ chosen via the Benjamini-Hochberg criterion (Section \ref{sec:error_rate_control})}
    \Output{A strongly causal graph $\widehat{\gcg}$}
    \Initialize{$S = [n]$  \texttt{\# unprocessed nodes}\\
      $E = \emptyset$  \texttt{\# edges of }$\widehat{\gcg}$\\
      $k = 1$ \texttt{\# a counter used only for notation}}

@@

\begin{remark}[Error Measurement]
  We we estimate the 1-step ahead prediction error by forming the variance matrix estimate

@@

\section{Application}
\label{sec:application}
In this section we apply our methods to a real set of EEG data
obtained from the ``EEG Database Data
Set''\footnote{http://archive.ics.uci.edu/ml/datasets/EEG+Database}
\cite{zhang1995event} on the UCI machine learning repository
\cite{uci_mlr}.  This dataset contains 1 second long measurements of
(64 channel) EEG signals from patients who are given visual stimuli
and are broken into two groups: control and alcoholic.  The dataset
then consists of
$\mathcal{D} = \{\big(x^{(i)}(t)\big)_{t = 1}^T, y^{(i)} \}_{i = 1}^N$
where $T = 256$, $x^{(i)}(t) \in \R^{64}$,
$y^{(i)} \in \{\text{A}, \text{C}\}$, and \todo{fill in N} $N = ????$.

We have constructed a simple pipeline for classifying each each trial
as either ``A'' or ``C'' by first applying an iterative PWGC algorithm
(see Algorithm \ref{alg:iterated_pwgc}) directly to the EEG data
$x^{(i)}$ to obtain a Granger-causality graph $\widehat{G}^{(i)}$ then
simply feeding the vectorized adjacency matrix of $\widehat{G}^{(i)}$
through a Kernel PCA stage and then a Support Vector Machine with
parameters fit by cross validation.

@@

Figures \ref{fig:ar_eeg_example} and \ref{fig:eeg_oos_eeror}
illustrate the appropriateness of modelling $x(t)$ with $\VAR$ models,
as opposed simply to a collection of $n$ unidimensional
autoregressiive models.  The conclusion being that the
Granger-causality graph estimates are not purely spurious.

@@

  {\scriptsize Figure \ref{fig:ar_eeg_example} provides an example of an
  autoregressive model fir via the Levinson-Durbin algorithm (choosing
  the system order via the BIC) on a single EEG channel.  Figure
  \ref{fig:eeg_oos_error} demonstrates the improvement of a single
  $\VAR$ model over independent $\mathsf{AR}$ models.

  Out-of-Sample estimates are performed on data that were not used in
  fitting the models.}

@@

\section{Conclusion}
\label{sec:conclusion}
In this paper we have argued that considering particular topological
properties of Granger-causality networks can provide substantial
theoretical insights as well as computational benefits.  In
particular, the notion of a strongly-causal graph has been exploited
to establish conditions under which pairwise causality testing alone
is sufficient for recovering a complete Granger-causality graph.
Moreover, examples from the literature suggest that such topological
assumptions may be reasonable in some applications and secondly, even
when the strong-causality assumption is not met, we have provided some
empirical evidence to suggest that our pairwise testing algorithm PWGC
can still outperform the LASSO and adaLASSO, both of which are
commonly employed in applications.  Application to more general graphs
may also be within reach via an iterative graph building process,
i.e. estimating a sequence of SCGs via PWGC applied to the residuals
of the previous model.

We emphasize that the causality graph topology is one of the key
defining features of time series analysis in comparison to standard
multivariate regression and therefore advocate for further study of
how different topological assumptions may impact the recovery of
causality graphs.  For example, are there provable guarantees on the
error rate of PWGC when applied to non strongly-causal graphs?
Computationally, can constraint systems or cunning adaptive weighting
schemes impose useful prior knowledge about graph topology for the
LASSO algorithm?  In another direction, it is known that the more
general notion of transfer entropy reduces to Granger-causality in the
case of Gaussian data \cite{barnett2009granger}, do our results extend
to causality networks defined via transfer entropy?  Finally, the work
of \cite{barnett2015granger} has established the superiority of
Granger-causality testing in state space models (as opposed to pure
autoregressions) in many cases.  Combining this work with our PWGC
algorithm (by modifying the approach described in Section
\ref{sec:pairwise_hypothesis_testing}) therefore is likely to enable
application to very large networks of time series data which are not
well approximated by finite $\VAR(p)$ models.

>>>>

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage[backend=bibtex,sorting=none]{biblatex}
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{fullpage}
\bibliography{\string~/Documents/academics/global_academics/global_bib.bib}
% \usepackage{apxproof}

@@

\begin{document}
\maketitle
\abstract{We study Granger-causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.
  We implement and detail finite-sample heuristics derived from our
  theory, and establish through simulation the efficiency gains (both
  statistical and computational) which can be obtained (in comparison
  to LASSO algorithms) when structural assumptions are met.  Finally,
  we provide an example application where we accurately discriminate
  between subjects in an EEG study based purely on the
  Granger-causality graphs inferred from data by our algorithms,
  demonstrating that meaningful features are captured by the
  Granger-causal graph topology.}

\keywords{Granger-causality, causality graph, time series, vector autoregression,
  LASSO, gene regulatory networks, network learning, EEG}

@@

\subsection{Contributions}
In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger-causality graph, which to our knowledge remains largely
unexplored.  The focus of this paper is to shed light on some of these
topological questions, in particular, we study a particularly simple
notion of causality graph topology which we term ``strongly causal''
and show that $\VAR$ models having this structure satisfy natural
intuitive notions of ``information flow'' through the graph.
Moreover, we show that such graphs are perfectly recoverable with only
\textit{pairwise} Granger-causality tests, which would otherwise suffer
from serious confounding problems.  Our finite sample results are
based on simulations, where we show promising results for these graph
topologies where our algorithm performs substantially better in our
simulation setup than do competing LASSO algorithms, even for graphs
that do not exactly satisfy our strongly-causal topology assumptions.

@@

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger-causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger-causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).
Secondly, we propose in section \ref{sec:structure_learning} a graph
search heuristic which implements our theoretical results to finite
data samples, specifying and summarizing appropriate methods for
hypothesis testing (section \ref{sec:pairwise_hypothesis_testing}),
model order selection (section \ref{sec:model_order_selection}),
computationally efficient estimation (section
\ref{sec:efficient_model_estimation}), and error rate controls
(section \ref{sec:error_rate_control}).  Our heuristics are compared
against the adaLASSO algorithm in Section
\ref{sec:empirical_evaluation}.  We stress the scalability of our
algorithm which is capable of comfortably handling hundreds or
thousands of nodes on a single machine, as opposed to standard LASSO
algorithms which do not take advantage of the special structure
associated with stationary time series data.  In section
\ref{sec:application} we develop an example application where a
classifier is constructed to accurately discriminate between subjects
in an EEG study based only on the Granger-causality graph topology
inferred by our algorithms.  Concluding remarks on further open
problems and extentions are provided in Section \ref{sec:conclusion}.

@@

    \BlankLine
    \caption{Finite Sample Pairwise Graph Recovery (PWGC)}
    \label{alg:finite_pwgc}

    \Input{Estimates of pairwise Granger-causality statistics $F_{ij}$
      (eqn \ref{eqn:gc_statistics}).  Matrix of edge probabilities $P_{ij}$ (eqn \ref{eqn:edge_inclusion_probability}).  Hypothesis testing threshold $\delta$ chosen via the Benjamini-Hochberg criterion (Section \ref{sec:error_rate_control})}
    \Output{A strongly causal graph $\widehat{\gcg}$}
    \Initialize{$S = [n]$  \texttt{\# unprocessed nodes}\\
      $E = \emptyset$  \texttt{\# edges of }$\widehat{\gcg}$\\
      $k = 1$ \texttt{\# a counter used only for notation}}

@@

\begin{remark}[Error Measurement]
  We estimate the 1-step ahead prediction error by forming the variance matrix estimate

@@

\section{Application}
\label{sec:application}
In this section we apply our methods to a real set of EEG data
obtained from the ``EEG Database Data
Set''\footnote{http://archive.ics.uci.edu/ml/datasets/EEG+Database}
\cite{zhang1995event} on the UCI machine learning repository
\cite{uci_mlr}.  This dataset contains 1 second long measurements of
(64 channel) EEG signals from patients who are given visual stimuli.
The subjects in the study are labeled as being either ``control'' or
``alcoholic'', however, we will ignore this label and instead focus on
distinguishing between subjects based on the Granger-causality graphs
inferred from the subject's trials.  Our reasoning is to focus on the
underlying question: ``Does our PWGC algorithm uncover meaningful
Granger-causal connections from EEG data?''.  Focusing only on the
subject label allows us to answer this question without simultaneously
grappeling with the physiological question of whether alcoholic
subjects as a group have discernable differences in their EEG
readings\footnote{Some exploratory analysis actually suggests
  that alcoholic Granger-causality graphs are not substantively
  different}, which is a stronger requirement than simply that there
are meaningful distinctions between the EEG readings of subjects
generally.

The dataset consists of
$\mathcal{D} = \{\big(x^{(i)}(t)\big)_{t = 1}^T, y^{(i)} \}_{i = 1}^N$
where $T = 256$, $x^{(i)}(t) \in \R^{64}$,
$y^{(i)} \in [N_{subjects}]$ (with $N_{subjects} = 119$), and
$N = 10723$.  There are on average 90 trials for each subject,
ranging between 30 and 119.

We have constructed a simple pipeline for discriminating between
subjects by first applying an iterative PWGC algorithm (see Algorithm
\ref{alg:iterated_pwgc}) directly to the EEG data $x^{(i)}$ to obtain
a Granger-causality graph $\widehat{G}^{(i)}$.  We then feed the
vectorized adjacency matrix of $\widehat{G}^{(i)}$ through a
polynomial Kernel multinomial logistic regression model with
parameters fit by cross validation.

The iterated application of PWGC is an example of a heuristic by which
graph estimates that aren't constrained to being strongly causal can
be obtained.  Theoretical analysis of more sophisticated heuristics
than PWGC (Algorithm \ref{alg:finite_pwgc}) is deferred to future work.

  \BlankLine
  \caption{Iterated PWGC Heuristic}
  \label{alg:iterated_pwgc}

  \Input{Data $x(t)$, maximum number of iterations $N$,
    threshold $r \in (0, 1)$.}
  \Output{Granger-causality graph estimate $\widehat{G}$ obtaind from
    iterated application of PWGC Algorithm \ref{alg:finite_pwgc}}
  \Initialize{$i = 1$, $\widehat{x}_0(t) \leftarrow x(t)$,
    $\sigma_0 \leftarrow Var\ x(t)$}

  \While{$i \le N$} {
    Estimate $\VAR$ model via PWGC on $x_{i - 1}(t)$ to obtain $\widehat{x}_{i - 1}(t)$ and $\widehat{G}_{i - 1}$\\
    $\epsilon_i(t) \leftarrow x_{i - 1}(t) - \widehat{x}_{i - 1}(t)$ \texttt{\# Compute residuals}\\
    $\sigma_i^2 \leftarrow \text{MSE}\big(\epsilon_i(t)\big)$ \texttt{\# Compute MSE}\\
    \If{$\sigma_i^2 > r \sigma_{i - 1}^2$}{
      \texttt{break  \# Insignificant Improvement}
    }
    $i \leftarrow i + 1$
  }
  $\widehat{G} \leftarrow \bigcup_{i} \widehat{G}_{i}$ \texttt{\# Combine PWGC graph estimates}\\
  \Return{$\widehat{G}$}
\end{algorithm}

@@

Figures \ref{fig:ar_eeg_example} and \ref{fig:eeg_oos_error}
illustrate the appropriateness of modelling $x(t)$ with $\VAR$ models,
as opposed simply to a collection of $n$ unidimensional
autoregressiive models.  The conclusion being that the
Granger-causality graph estimates are not purely spurious.

@@

  {\scriptsize Figure \ref{fig:ar_eeg_example} provides an example of
    an autoregressive model fir via the Levinson-Durbin algorithm
    (choosing the system order via the BIC) on a single EEG channel.
    Figure \ref{fig:eeg_oos_error} demonstrates the improvement of a
    single $\VAR$ model over independent $\mathsf{AR}$ models,
    providing evidence that intra-node edges inferred by PWGC are not
    simply spurious.

  Out-of-Sample estimates are performed on data that were not used in
  fitting the models.}
\end{figure}

We provide the final results in Figure
\ref{fig:logistic_regression_results} where the classification
accuracy is quantified by the (multiclass) MCC of $\approx 0.20$ on a
held out validation set consisting of $20\%$ of the data.  This MCC
score exceeds by a modest margin the performance of randomly guessing,
lending strong evidence to the assertion that PWGC successfully
recovers some meaningful differences between the Granger-causality
graphs of different subjects, particularly if we recall that there are
generally fewer examples for each subject than there are subjects
overall.  Moreover, distinguishing only between two particular
subjects (as opposed to distinguishing between one subject and $118$
other possibilities) is substantially easier -- an illustration is
provided in Figure \ref{fig:logistic_regression_2pair} where we embed
the data into the plane through supervised dimensionality reduction.
In general, fitting a classifier between two subjects achieves an
$MCC$ in excess of $0.6$ and up to $0.9$.

\begin{figure}
  \centering
  \caption{Subject Classification from Granger-causality Graphs}
  \begin{subfigure}[b]{0.45\textwidth}
    \caption{Distinguishing Individual Subjects}
    \label{fig:logistic_regression_2pair}
    \includegraphics[width=\linewidth]{logistic_regression_separation.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \caption{Complete Multiclass Classifier}
    \label{fig:logistic_regression_results}
    \includegraphics[width=\linewidth]{logistic_regression_confusion_matrix.pdf}

    {\scriptsize }
  \end{subfigure}

  {\scriptsize Figure \ref{fig:logistic_regression_2pair} illustrates
    discriminating between two particular subjects based on their
    EEG Granger-causality graphs.  In this case $MCC = 0.75$ on held
    out data.  Visualization is constructed by supervised
    dimensionality reduction and is purely illustrative.

    Figure \ref{fig:logistic_regression_results} provides the
    row-normalized confusion matrix (computed on held out validation
    data) of a multiclass logistic regression classifier used to
    classify subjects based on their EEG Granger-causality graphs.  $MCC = 0.20$}

@@

\section{Conclusion}
\label{sec:conclusion}
In this paper we have argued that considering particular topological
properties of Granger-causality networks can provide substantial
theoretical insights as well as computational benefits.  In
particular, the notion of a strongly-causal graph has been exploited
to establish conditions under which pairwise causality testing alone
is sufficient for recovering a complete Granger-causality graph.
Moreover, examples from the literature suggest that such topological
assumptions may be reasonable in some applications and secondly, even
when the strong-causality assumption is not met, we have provided
simulation evidence to suggest that our pairwise testing algorithm
PWGC can still outperform the LASSO and adaLASSO, both of which are
commonly employed in applications.  Application to inference of more
general graphs may also be within reach via an iterative graph
building process as suggested in Section \ref{sec:application}.

Our application Section \ref{sec:application} provides strong evidence
that PWGC is capable of uncovering meaningful features from networks
of time series data by constructing a classifier which accurately
discriminates between subjects in an EEG experiment based purely on
the Granger-causality graph topology inferred by the iterated
application of PWGC.

We emphasize that the causality graph topology is one of the key
defining features of time series analysis in comparison to standard
multivariate regression and therefore advocate for further study of
how different topological assumptions may impact the recovery of
causality graphs.  For example, are there provable guarantees on the
error rate of PWGC when applied to non strongly-causal graphs?
Computationally, can constraint systems or cunning adaptive weighting
schemes impose useful prior knowledge about graph topology for the
LASSO algorithm?  In another direction, it is known that the more
general notion of transfer entropy reduces to Granger-causality in the
case of Gaussian data \cite{barnett2009granger}, do our results extend
to causality networks defined via transfer entropy?  Finally, the work
of \cite{barnett2015granger} has established the superiority of
Granger-causality testing in state space models (as opposed to pure
autoregressions) in many cases.  Combining this work with our PWGC
algorithm (by modifying the approach described in Section
\ref{sec:pairwise_hypothesis_testing} to instead utilize state-space
Granger-causality testing) therefore is likely to enable application
to very large networks of time series data which are not well
approximated by finite $\VAR(p)$ models.
